---
title: "locFDR"
author: "Redgar"
date: "March 29, 2017"
output: html_document
---

Local FDR for identification of outliers

# Hannum 2013 mol.cell
Methylation quality control
We used principal component (PC) analysis to identify and remove outlier samples. We
converted each sample into a z-score statistic, based on the squared distance of its 1st PC
from the population mean. The z-statistic was converted to a false-discovery rate using the
Gaussian cumulative distribution and the Benjamini-Hochberg procedure (Benjamini and
Hochberg, 1995). Samples falling below an FDR of 0.2 were designated at outliers and
removed. This filtering procedure was performed iteratively until no samples were
determined to be an outlier. A total of 24 samples were removed in this manner.


The PCA to run locfdr on
```{r}
library(locfdr)


### PCA
Beta_sample_clean<-EL_CG_Clean_mnImputed 

PCA<-as.data.frame(unclass(princomp(Beta_sample_clean)$loadings))

# Match to meta data
meta_ordered<-meta[which(meta$Sample_Name%in%colnames(Beta_sample_clean)),]
meta_ordered<-meta_ordered[match(colnames(Beta_sample_clean), meta_ordered$Sample_Name),]

# 2nd comp sex (mind the mix ups)
ggplot(PCA,aes(Comp.1,Comp.2, size=meta_ordered$Dataset, color=meta_ordered$gender))+geom_point(shape=19)+theme_bw()+scale_size_manual(values=c(3,5))

# Hannum method remove PCA based outliers
zstat_comp1<-((PCA$Comp.1-mean(PCA$Comp.1))/sd(PCA$Comp.1))

w <- locfdr(zstat_comp1)
w$fp0

which(w$fdr < 0.2)## good sample line (kinda random threshold)

    # Don't remember what these lines do
    # sum(which(w$fdr < 0.2) > 200)/sum(w$fdr < 0.2)
    # w$mat[which(w$mat[, "fdr"] < 0.2)[1], "Fdrright"]

outmeta<-meta_ordered[which(w$fdr < 0.2),]
outmeta$PCA<-"FDR<0.2"
inmeta<-meta_ordered[which(!(w$fdr < 0.2)),]
inmeta$PCA<-"FDR>0.2"

outliers<-as.character(outmeta$Sample_Name)

meta_out<-rbind(outmeta,inmeta)
meta_out<-meta_out[match(colnames(Beta_sample_clean), meta_out$Sample_Name),]

# Plot PCA colored by whether they were considered an outlier by locFDR
ggplot(PCA,aes(Comp.1,Comp.2, color=meta_out$PCA))+geom_point(shape=19, size=3)+theme_bw()+scale_size_manual(values=c(3,5))
```

Advised to run locFDR itteratively until there are no outliers detected, in EL_CG only one round of outlier removal was necessary. By the second round no outliers were detected
```{r}
## remove outliers and rerun PCA
Beta_sample_clean_outrm<-EL_CG_Clean_mnImputed[,which(!(colnames(EL_CG_Clean_mnImputed)%in%outliers))]

## Repeat same code to see if any outliers called this time
PCA2<-as.data.frame(unclass(princomp(Beta_sample_clean_outrm)$loadings))
meta_ordered<-meta[which(meta$Sample_Name%in%colnames(Beta_sample_clean_outrm)),]
meta_ordered<-meta_ordered[match(colnames(Beta_sample_clean_outrm), meta_ordered$Sample_Name),]
zstat_comp1<-((PCA2$Comp.1-mean(PCA2$Comp.1))/sd(PCA2$Comp.1))
w <- locfdr(zstat_comp1)
w$fp0
which(w$fdr < 0.2)## good line
sum(which(w$fdr < 0.2) > 200)/sum(w$fdr < 0.2)
w$mat[which(w$mat[, "fdr"] < 0.2)[1], "Fdrright"]
outmeta<-meta_ordered[which(w$fdr < 0.2),]
outmeta$PCA<-"FDR<0.2"
inmeta<-meta_ordered[which(!(w$fdr < 0.2)),]
inmeta$PCA<-"FDR>0.2"
meta_out<-rbind(outmeta,inmeta)
meta_out<-meta_out[match(colnames(Beta_sample_clean_outrm), meta_out$Sample_Name),]
ggplot(PCA,aes(Comp.1,Comp.2, color=meta_out$PCA))+geom_point(shape=19, size=3)+theme_bw()+scale_size_manual(values=c(3,5))


# 3rd comp?
ggplot(PCA,aes(Comp.2,Comp.3, size=meta_ordered$Dataset, color=meta_ordered$ethnicity.bin))+geom_point(shape=19)+theme_bw()+scale_size_manual(values=c(3,5))

# 4th comp ethnicity
ggplot(PCA,aes(Comp.3,Comp.4, size=meta_ordered$Dataset, color=meta_ordered$ethnicity.bin))+geom_point(shape=19)+theme_bw()+scale_size_manual(values=c(3,5))

# 5th comp?
ggplot(PCA,aes(Comp.4,Comp.5, size=meta_ordered$Dataset, color=meta_ordered$age))+geom_point(shape=19)+theme_bw()+scale_size_manual(values=c(3,5))
```


