---
title: "Kollman_Vaccination"
author: "Rachel Edgar"
date: "08/05/2017"
output: html_document
---

```{r setup, include=FALSE}
setwd("~/Side/Kollman_Vaccination")

library(methylumi)
library(wateRmelon)
library(RPMM)
library("ggsci")
library(reshape)
library(ggplot2)
library(RColorBrewer)
library(lme4)
library(gridExtra)

```


## Load the data from Genome Studio and make methylumi objects

```{r,eval=F, echo=T}
allFile <- ("/home/redgar/Side/Kollman_Vaccination/Kollman_Pilot-alldata.txt") 
qcFile <- ("/home/redgar/Side/Kollman_Vaccination/Kollman_Pilot-qcfile.txt") 

#vaccination<- lumiMethyR(allFile)
vaccination.2 <- methylumiR(allFile, qcfile = qcFile)
save(vaccination.2, file="Kollman_Vaccination.RData")
```


```{r, echo=F}
load("Kollman_Vaccination.RData")
dim(vaccination.2)#865918       31 
```

## Meta data
```{r}
meta_shipping<-read.csv("/home/redgar/Side/Kollman_Vaccination/Shipping_info_WB_epigenetics_Kobor_lab.csv")
meta_sampleinfo<-read.table("/home/redgar/Side/Kollman_Vaccination/Kollman_Pilot-samplefile.txt", sep="\t", header=T, stringsAsFactors = F)
meta_sampleinfo$Sample_ID[24]<-"GR02_Visit6"

meta_shipping<-meta_shipping[-31,]
meta_shipping$Sample_ID<-gsub(" ","",meta_shipping$VisitNum)
meta_shipping$Sample_ID<-paste(meta_shipping$SUBJECT_ID, "_",meta_shipping$Sample_ID, sep="")

meta_vaccination<-merge(meta_shipping, meta_sampleinfo, by="Sample_ID")
meta_vaccination$Sample_ID<-meta_vaccination$Sample_Label

meta_vaccination<-meta_vaccination[ , -which(names(meta_vaccination) %in% c("Sample_Label","Sample_Group","ShippedTo","ShippedFrom",
                                                                        "ShippedDate","Container","NumberShipped","SampleYN",
                                                                        "Cells.per.vial","SampleYN","test","DeviationYN","Deviation","Pool_ID"))]
meta_vaccination<-meta_vaccination[match(sampleNames(vaccination.2), meta_vaccination$Sample_ID),]

dim(meta_vaccination)
identical(sampleNames(vaccination.2), as.character(meta_vaccination$Sample_ID))

save(meta_vaccination, file="meta_vaccination.RData")
```

#### Pvalues of samples on each array
```{r}
avgPval <- colMeans(pvals(vaccination.2))

#p-values for detection for the samples
meta_vaccination$Det_pval<-avgPval

ggplot(meta_vaccination)+geom_boxplot(aes(as.factor(Sentrix_ID), Det_pval, fill=as.factor(Sentrix_ID)), outlier.size=NA)+
  geom_point(aes(as.factor(Sentrix_ID), Det_pval, group=Sample_ID, fill=as.factor(Sentrix_ID)), shape=21, color="black",
             position = position_jitter(w = 0.25))+theme_bw()
```


#### Beta distribution
```{r}
vaccinationBetas<-betas(vaccination.2)

## Beta distribtuions 
Beta_melted<- melt(vaccinationBetas)
#remove NAs before plotting (otherwise get many non-inifnite warnings)
Beta_Plot<-Beta_melted[which(!(is.na(Beta_melted$value))),]

#add meta
Beta_Plot<-merge(Beta_Plot,meta_vaccination, by.x="X2", by.y="Sample_ID")

ggplot(Beta_Plot, aes(value, group=X2, color=SID))+
  geom_density()+theme_bw()+scale_color_locuszoom()+xlab("DNAm Beta Value")

```


#### Confirm ID with SNPs and cluster by DNAm type?
```{r}
# Clustering By Any Meta Data variable
# remove rows with NAs
vaccinationBetasbeta_cluster<-vaccinationBetas[complete.cases(vaccinationBetas),]

# plot clustering with color function
plotHclustColors <- function(matrix,leafcolor) {
  colnames(matrix) <- leafcolor
  d <- dist(t(matrix))
  hc <- hclust(d, method = "complete") #single, complete, average, ward
  color<-rep(brewer.pal(12,"Paired"),ceiling(length(unique(leafcolor))/12))
  labelColors <- color[sample(1:length(color),length(unique(leafcolor)))]
  colLab <- function(n) {
    if (is.leaf(n)) {
      a <- attributes(n)
      labCol <- labelColors[which(unique(leafcolor) == a$label)]
      attr(n, "nodePar") <- c(a$nodePar, lab.col=labCol)
    }
    n
  }
  clusDendro <- dendrapply(as.dendrogram(hc), colLab)
  plot(clusDendro)
}


# Plot
plotHclustColors(vaccinationBetasbeta_cluster, meta_vaccination$SID) 


#Genotyping Probes (show that samples from the same individual cluster by SNPs)
CpGs<- rownames(vaccinationBetas)
SNP_Probes<-CpGs[grep("rs", CpGs)]# 59 SNP probes on EPIC
SNPs<-vaccinationBetas[CpGs%in%SNP_Probes,]
# Remove NAs (if any)
SNPs<-SNPs[complete.cases(SNPs),]# 58 cause one was all NA

plotHclustColors(SNPs, meta_vaccination$SID)#all samples pair with their other time point
```


# Probe Filtering
##### Removal of SNP Probes
We remove the SNP probes as they are used as an internal control to ensure your samples are what you think they are and are not used for any methylation analysis.

```{r, echo=FALSE}
vaccination.2_filtered <- vaccination.2[substring(featureNames(vaccination.2),1,2) != "rs", ]
dim(vaccination.2_filtered) # probes = 865859, n = 31
```


##### Cross-hybridizing probes and polymorphic probes. Don't have an a Magda Price annotation for EPIC. So gonna go find that shit. 
https://genomebiology.biomedcentral.com/articles/10.1186/s13059-016-1066-1
"43,254 cross-reactive probes with ≥ 47 bp homology with an off-target site, of which 15,782 (36.5 %) are new to the EPIC platform"

They include this annotated list in their supplement.

```{r}
cross_reactive<-read.csv("~/Side/5hmC_Turecki/Pidsley_cross_reactive.csv", stringsAsFactors = F)
vaccination.2_filtered<-vaccination.2_filtered[which(!(featureNames(vaccination.2_filtered)%in%cross_reactive$X)),]
dim(vaccination.2_filtered) # probes = 822682, n = 31, 43177 filtered
```

For polymorphic probes I will The Pidsley annotation aswell for "Probes overlapping genetic variants at targeted CpG sites." and "Probes overlapping genetic variants at single base extension sites for Infinium Type I probes" but NOT "Probes with genetic variants overlapping the body of the probe: 48 base pairs for Infinium Type I probes and 49 base pairs for Infinium Type II probes."

```{r}
polymorphic<-read.csv("~/Side/5hmC_Turecki/Pidsley_Polymorphic_CpGs.csv",stringsAsFactors = F)
length(unique(polymorphic$PROBE))
baseext<-read.csv("~/Side/5hmC_Turecki/Pidsley_single_base_extension.csv",stringsAsFactors = F)
length(unique(baseext$PROBE))

vaccination.2_filtered<-vaccination.2_filtered[which(!(featureNames(vaccination.2_filtered)%in%c(polymorphic$PROBE, baseext$PROBE))),]
dim(vaccination.2_filtered) # probes = 811063, n = 31, 11619 filtered
```

##### Sex Chromosomes need to add
```{r}
vaccination.2_filtered <- vaccination.2_filtered[!featureData(vaccination.2_filtered)$CHR%in%c("X", "Y"), ]
dim(vaccination.2_filtered) # probes = 792939, n = 31, 18124 filtered
```


We have removed 72,979 probes. This leaves us with 792,939 probes for our analysis. 



# Watermelon Filter and Normalization
# BMIQ (probe type normalization)
    #Checks
      #iDMR, X, Snps
      #Replicates correlation
      
```{r,eval=F, echo=T}
# waterMelon Bad probe filtration
vaccination.pf<-pfilter(vaccination.2_filtered)

    # 0 samples having 1 % of sites with a detection p-value greater than 0.05 were removed 
    # Samples removed:  
    # 1067 sites were removed as beadcount <3 in 5 % of samples 
    # 669 sites having 1 % of samples with a detection p-value greater than 0.05 were removed 
dim(vaccination.pf) #  791,273       31 

vaccination.2_filtered<-vaccination.pf
save(vaccination.2_filtered, file = "vaccination_fully_filtered.RData")
```

We have removed another 1,666 probes. This leaves us with 791,273 probes for our analysis. 

#### Probe attrition plot
```{r}
df<-data.frame(sample_num_remaining=c(865918,865859,822682,811063,792939,791872,791273), filter=c("EPIC Probe Number","Removal of SNP Probes","Removal of Pidsley Cross Reactive Probes","Removal of Pidsley Polymorphic Probes","Removal of X and Y chromosome probes", "Removal of Probes with Beadcount <3\nin 5 % of Samples", "Removal of Probes with 1 % of samples\nwith a detection p-value greater than 0.05"))
df$sample_num_lost<-c(0,sapply(2:nrow(df), function(x) df$sample_num_remaining[x-1]-df$sample_num_remaining[x]))

df$filter<-factor(df$filter, rev(df$filter))
          
library(scales)
ggplot(df)+
  geom_bar(aes(filter,-sample_num_remaining), stat="identity", fill="grey70", color="black")+
  geom_bar(aes(filter,sample_num_lost), stat="identity",fill="darkred", color="black")+
  geom_text(aes(x=filter, y=-min(sample_num_remaining)/2,  label=comma(sample_num_remaining)))+
  geom_text(aes(x=filter, y=max(sample_num_lost)/1.5,  label=comma(sample_num_lost)))+
  geom_hline(yintercept=0)+
  coord_flip()+theme_bw()+ylab("")+xlab("")+
  theme(axis.line = element_blank(),
        axis.ticks = element_blank(),
        axis.text.x = element_blank(),
        axis.text.y = element_text(colour = "grey20", size=12),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) +
  scale_x_discrete(position = "top") 
```



# Normalization 
```{r}
# dasen
vaccination_dasen<-dasen(vaccination.2_filtered) 

## Beta distribtuions plot
Beta_melted<- melt(betas(vaccination_dasen))
Beta_Plot<-Beta_melted[which(!(is.na(Beta_melted$value))),]
Beta_Plot<-merge(Beta_Plot,meta_vaccination, by.x="X2", by.y="Sample_ID")

ggplot(Beta_Plot, aes(value, group=X2, color=SID))+
  geom_density()+theme_bw()+scale_color_locuszoom()+xlab("DNAm Beta Value")


dim(vaccination_dasen) # 791273      31 
vaccinationBetas<-betas(vaccination_dasen)
vaccinationBetas<-as.data.frame(vaccinationBetas)
save(vaccinationBetas, file = "vaccination_prbFilter_dasen_norm.RData")

```




## Blood composition correction
Don't load packages for this section, somthing is overwriting libraries for ECC
```{r}
library("FlowSorted.Blood.450k")
source("~/ECC5.R")

load("vaccination_prbFilter_dasen_norm.RData")

predictions<- ECC5(vaccinationBetas)
counts<- as.data.frame(predictions$counts)

save(counts, file="Vaccinations_cell_compositions.RData")


# Impute NAs
imputeMedianv3<-function(x) apply(x, 1, function(x){x[is.na(x)]<-median(x, na.rm=T); x}) #impute with row mean
vaccination_betas_imputed<-t(imputeMedianv3(vaccinationBetas))

# deconvolute
avebeta.lm<-apply(vaccination_betas_imputed, 1, function(x){
  counts[colnames(vaccination_betas_imputed),]->blood
  lm(x~CD8T+CD4T+NK+Bcell+Mono,data=blood)
})
residuals<-t(sapply(avebeta.lm, function(x)residuals(summary(x))))
colnames(residuals)<-colnames(vaccination_betas_imputed)
adj.residuals<-residuals+matrix(apply(vaccination_betas_imputed, 1, mean), nrow=nrow(residuals), ncol=ncol(residuals))


save(adj.residuals, file="Vaccinations_adjusted_betas.RData")


# do composition again (so can see the normalization effect after adjustment)
load("Vaccinations_adjusted_betas.RData")
predictions<- ECC5(as.data.frame(adj.residuals))
counts_adjusted<- as.data.frame(predictions$counts)
save(counts_adjusted, file="Vaccinations_cell_compositions_adjusted.RData")


#Blood plot
load("Vaccinations_cell_compositions.RData")
load("Vaccinations_cell_compositions_adjusted.RData")

counts$Sample_ID<-rownames(counts)
cellprop_melt<-melt(counts, id="Sample_ID")
cellprop_melt$Data<-"Original Predicated Proportions"
counts_adjusted$Sample_ID<-rownames(counts_adjusted)
cellprop_melt_adjusted<-melt(counts_adjusted, id="Sample_ID")
cellprop_melt_adjusted$Data<-"Proportions After Deconvolution"
cellprop_melt_blood<-rbind(cellprop_melt, cellprop_melt_adjusted)
cellprop_melt_blood$Data <- factor(cellprop_melt_blood$Data,c("Original Predicated Proportions", "Proportions After Deconvolution"))

ggplot(cellprop_melt_blood, aes(variable, value, fill=as.factor(variable)))+
  geom_point(shape=21, color="black", position = position_jitter(w = 0.2))+theme_bw()+xlab("Blood Cell Type")+ylab("Cell Type Proportion")+
  facet_grid(~Data, scales="free_x")+scale_fill_manual(values=c("#fed976", "#feb24c", "#fd8d3c", "#fc4e2a", "#e31a1c", "#bd0026"),guide=FALSE)
```

## blood compostion association with variables
```{r}
levels(meta_vaccination$VisitNum)
meta_vaccination$visitnumeric<-meta_vaccination$VisitNum
levels(meta_vaccination$visitnumeric)
levels(meta_vaccination$visitnumeric)<-c("","3","4","5","6","7")
meta_vaccination$visitnumeric<-as.numeric(as.character(meta_vaccination$visitnumeric))

meta_vaccination<-meta_vaccination[match(rownames(counts), meta_vaccination$Sample_ID),]

library(lme4)
cell_comp_pvalues<-sapply(1:6, function(x){
  modl<-lmer(counts[,x] ~ meta_vaccination$visitnumeric + (1|meta_vaccination$SID))
  mods<-lmer(counts[,x] ~  (1|meta_vaccination$SID))
  Likelihood_Ratio_Test<-anova(modl, mods)
  Likelihood_Ratio_Test[["Pr(>Chisq)"]][2]
})

p.adjust(cell_comp_pvalues, method = "bonferroni", n = length(cell_comp_pvalues))

counts$Sample_ID<-rownames(counts)
counts_plot<-melt(counts)
plt<-merge(counts_plot, meta_vaccination, by="Sample_ID")
ggplot(plt, aes(visitnumeric, value, group=SID, color=SID))+geom_line()+geom_point()+theme_bw()+facet_wrap(~variable)+scale_color_locuszoom()+xlab("Visit")+ylab("Cell Proportion in Whole Blood Sample")

```



## Heat scree plot
```{r}
source("Heat_scree_plot.R")

load("meta_vaccination.RData")
load("Vaccinations_adjusted_betas.RData")

vaccionationBetas<-as.data.frame(adj.residuals)

# PCA
PCA_full<-prcomp(na.omit(vaccionationBetas))
Loadings<-as.data.frame(unclass(PCA_full$rotation))
vars <- PCA_full$sdev^2
Importance<-vars/sum(vars)



#Restructure meta
meta_vaccination$Sentrix_ID<-as.factor(meta_vaccination$Sentrix_ID)

meta_vaccination$visitnumeric<-meta_vaccination$VisitNum
levels(meta_vaccination$visitnumeric)<-c("","3","4","5","6","7")
meta_vaccination$visitnumeric<-as.numeric(as.character(meta_vaccination$visitnumeric))


meta_categorical <- meta_vaccination[, c("SID","Sex","VisitNum","CollectionDate","Sentrix_ID")]  # input column numbers in meta that contain categorical variables
meta_continuous <- meta_vaccination[, c("Age.yrs.","visitnumeric")]  # input column numbers in meta that contain continuous variables
colnames(meta_categorical) <- c("Subject ID","Sex","Visit Number","Collection Date","Sentrix_ID")
colnames(meta_continuous) <- c("Age","Visit Numeric")

ord<-1:length(c(colnames(meta_categorical),colnames(meta_continuous)))

# how far do you want the plot to go?
PCs_to_view<-10


heat_scree_plot(Loadings, Importance)
```

## ComBat
```{r}
library(sva)
library(pamr)
library(limma)
Mval<-function(beta) log2(beta/(1-beta))

pheno = meta_vaccination
edata = apply(vaccionationBetas, 1, Mval) # need mvalues for combat
edata = as.data.frame(edata)
edata = t(edata)


pheno$Sentrix_ID<-as.factor(pheno$Sentrix_ID)

#Combat

#Sentrix_ID
batch = pheno$Sentrix_ID
combat_vaccionationBetas = ComBat(dat=edata, batch=batch, mod=NULL, par.prior=TRUE)

#another heat scree
# PCA
PCA_full<-prcomp(na.omit(combat_vaccionationBetas))
Loadings<-as.data.frame(unclass(PCA_full$rotation))
vars <- PCA_full$sdev^2
Importance<-vars/sum(vars)
heat_scree_plot(Loadings, Importance)


#Back to betas
betas<-function(M) 2^M/((2^M)+1)
combat_vaccionationBetas2 = apply(combat_vaccionationBetas, 1, betas) # need mvalues for combat
combat_vaccionationBetas2 = as.data.frame(combat_vaccionationBetas2)
combat_vaccionationBetas2 = t(combat_vaccionationBetas2)
combat_vaccionationBetas<-combat_vaccionationBetas2

save(combat_vaccionationBetas, file="Vaccination_betas_dasen_cellcomp_combat.RData")
```



## Replicates
One replicate so confirm it has a lower RMSE with processing
```{r}
load("Kollman_Vaccination.RData")
raw_betas<-betas(vaccination.2)
raw_betas_GR02_Visit6<-raw_betas[,grep("GR02_Visit6",colnames(raw_betas))]

load("vaccination_prbFilter_dasen_norm.RData")
dasen_betas_GR02_Visit6<-vaccinationBetas[,grep("GR02_Visit6",colnames(vaccinationBetas))]

load("Vaccinations_adjusted_betas.RData")
cellcomp_betas_GR02_Visit6<-as.data.frame(adj.residuals[,grep("GR02_Visit6",colnames(adj.residuals))])

load("Vaccination_betas_dasen_cellcomp_combat.RData")
combatted_betas_GR02_Visit6<-combat_vaccionationBetas[,grep("GR02_Visit6",colnames(combat_vaccionationBetas))]



## RMSE
library(hydroGOF)
rep_cor<-data.frame(Comparison=c("Original", "dasen Normalized",  
                                 "dasen Normalized \nCell Composition Adjusted", 
                                 "dasen Normalized \nCell Composition Adjusted\n ComBatted"),
                    Correlation=c(rmse(raw_betas_GR02_Visit6[,1],raw_betas_GR02_Visit6[,2], use="complete.obs"),
                                  rmse(dasen_betas_GR02_Visit6[,1],dasen_betas_GR02_Visit6[,2], use="complete.obs"),
                                  rmse(cellcomp_betas_GR02_Visit6[,1],cellcomp_betas_GR02_Visit6[,2], use="complete.obs"),
                                  rmse(combatted_betas_GR02_Visit6[,1],combatted_betas_GR02_Visit6[,2], use="complete.obs")))


rep_cor$Comparison<-factor(rep_cor$Comparison, levels =c("Original", "dasen Normalized", 
                                                         "dasen Normalized \nCell Composition Adjusted", 
                                                         "dasen Normalized \nCell Composition Adjusted\n ComBatted"))
ggplot(rep_cor, aes(Comparison, Correlation))+geom_point(size=2)+theme_bw()+ylab("Root Mean Square Error")+xlab("")

# remove replicate

vaccionationBetas<-combat_vaccionationBetas[,-grep("GR02_Visit6_rep",colnames(combat_vaccionationBetas))]
meta_vaccination<-meta_vaccination[which(meta_vaccination$Sample_ID%in%colnames(vaccionationBetas)),]
meta_vaccination<-meta_vaccination[match(colnames(vaccionationBetas),meta_vaccination$Sample_ID),]

save(meta_vaccination, vaccionationBetas, file="vaccination_cleaned_data.RData")
```


# Remove Blood invariable CpGs
```{r}
library(RCurl)
x <- getURL("https://raw.githubusercontent.com/redgar598/Tissue_Invariable_450K_CpGs/master/Invariant_Blood_CpGs.csv")
y <- read.csv(text = x)

vacc_independent_blood_invariable<-vaccionationBetas[which(rownames(vaccionationBetas)%in%y$CpG),]#102173/120009 of the independnt invariable sites are in PAWS

# Call varibility in PAWS
Variation<-function(x) {quantile(x, c(0.9), na.rm=T)[[1]]-quantile(x, c(0.1), na.rm=T)[[1]]}
vacc_ref_range<-sapply(1:nrow(vacc_independent_blood_invariable), function(x) Variation(vacc_independent_blood_invariable[x,]))
Invariable_in_vacc<-vacc_independent_blood_invariable[which(vacc_ref_range<0.05),]

# Which CpGs are invariable in PAWS and the independent data
invar_in_vacc_and_independent<-intersect(y$CpG, rownames(Invariable_in_vacc)) #101869/102173 (99.7%)
vaccionationBetas_variable<-vaccionationBetas[which(!(rownames(vaccionationBetas)%in%invar_in_vacc_and_independent)),]#689404 
vaccionationBetas<-vaccionationBetas_variable #689404 vs 791273 sites

save(vaccionationBetas, meta_vaccination, file="vaccination_cleaned_data_variable.RData")
```


## Linear Mixed effects One way-ANOVA
```{r}
load("vaccination_cleaned_data_variable.RData")
Mval<-function(beta) log2(beta/(1-beta))
vaccionationMval = apply(vaccionationBetas, 1, Mval) # need mvalues for combat
vaccionationMval = as.data.frame(vaccionationMval)
vaccionationMval = t(vaccionationMval)


### Linear Mixed effects model
visit_pvalues<-sapply(1:nrow(vaccionationMval), function(x){
  print(x)
  meta_ex<-meta_vaccination
  meta_ex$Mval<-unlist(vaccionationMval[x,])
  modl<-lmer(Mval ~ as.factor(VisitNum) + (1|SID), data=meta_ex)
  mods<-lmer(Mval ~  (1|SID), data=meta_ex)
  Likelihood_Ratio_Test<-anova(modl, mods)
  Likelihood_Ratio_Test[["Pr(>Chisq)"]][2]
})

save(visit_pvalues, file="vaccination_visit_pvalues.RData")


## pvalue distribution
load("vaccination_visit_pvalues.RData")
pvalue_dist<-data.frame(CpG=rownames(vaccionationBetas), Nominal_P=visit_pvalues)
ggplot(pvalue_dist, aes(Nominal_P))+geom_histogram(fill="grey90", color="black")+theme_bw()+xlab("Nominal P Value")

## multiple test correction
bon_pval<-p.adjust(visit_pvalues, method = "bonferroni", n = length(visit_pvalues))
fdr_pval<-p.adjust(visit_pvalues, method = "fdr", n = length(visit_pvalues))


stat_hits<-vaccionationBetas[which(fdr_pval<0.1),]
dim(stat_hits)

```


## figure out a delta beta
not sure how to decompose the random effects over the visits, so I am just going to calcuate a paired delta beta
if a CpG has an overall sig effect of visit than can ask if they also have a large delta beta between any visits

Going to test all CpGs so have deltabetas for volcano plot
```{r}
#Visitone compared to each visit per person
visit_db<-lapply(1:nrow(vaccionationBetas), function(x){
  meta_ex<-meta_vaccination
  meta_ex$Beta<-unlist(vaccionationBetas[x,])

  meta_ex$SID<-as.character(meta_ex$SID)
  meta_ex$VisitNum<-as.character(meta_ex$VisitNum)
  
  mean_Betas<-tapply(meta_ex$Beta, list(meta_ex$VisitNum, meta_ex$SID), mean)
  
  colMeans(data.frame(Visit4dB=mean_Betas[1,]-mean_Betas[2,],
             Visit5dB=mean_Betas[1,]-mean_Betas[3,],
             Visit6dB=mean_Betas[1,]-mean_Betas[4,],
             Visit7dB=mean_Betas[1,]-mean_Betas[5,]))
})
visit_db<-do.call(rbind, visit_db) 
rownames(visit_db)<-rownames(stat_hits)

save(visit_db, file="delta_beta_visit.RData")

## max delta beta
load("delta_beta_visit.RData")
max_deltabeta<-sapply(1:nrow(visit_db), function(x) {
  maxbd<-visit_db[x,][which(abs(visit_db[x,])==max(abs(visit_db[x,]),na.rm=T))]
  if(length(maxbd)==0){NA}else{maxbd}
  })

Hit_info<-data.frame(CpG=rownames(vaccionationBetas), Nominal_P=visit_pvalues, fdr_pval=fdr_pval, max_deltabeta=max_deltabeta)

stat_biohits<-Hit_info[which(abs(Hit_info$max_deltabeta)>0.05 & Hit_info$fdr_pval<=0.1),]
```

# volcano
```{r}
source("volcano.R")
makeVolcano(fdr_pval, max_deltabeta, 0.05, 0.1, "DNAm Change from Initial Visit", 0.4)

```


# plot
```{r}
load("~/MethylationEPIC_v-1-0_B3.RData")
meta_vaccination$Visit <- factor(meta_vaccination$Visit,c("","Day 0","Day 1 post dose 1","Day 3 post dose 1",
                                                          "Day 7 post dose 1","Day 14 post dose 1"
                                                          ))
levels(meta_vaccination$Visit)<-c("","Day 0","Day 1","Day 3",
                                                          "Day 7","Day 14")


plt<-function(CpG){
  Betas<-vaccionationBetas[which(rownames(vaccionationBetas)%in%CpG),]
  Betas<-melt(Betas)
  colnames(Betas)<-c("CpG","Sample_ID","Beta")
  plt<-merge(Betas, meta_vaccination, by="Sample_ID")
  
  annotation_plt<-annotation_EPIC[which(annotation_EPIC$IlmnID%in%CpG),c("IlmnID","UCSC_RefGene_Name")]
  uni_gene<-sapply(strsplit(as.character(annotation_plt$UCSC_RefGene_Name),";"), function(x) paste(unique(x), sep=" "))
  annotation_plt$uni_gene<-sapply(1:length(uni_gene), function(x) paste0(uni_gene[[x]], collapse = " "))
annotation_plt$label<-paste(annotation_plt$IlmnID, annotation_plt$uni_gene, sep = " - ")
  
  plt<-merge(plt, annotation_plt, by.x="CpG", by.y="IlmnID")
  ggplot(plt, aes(Visit, Beta, group=SID, color=SID))+geom_line()+geom_point()+theme_bw()+facet_wrap(~label)+scale_color_locuszoom()+xlab("Visit")+ylab("DNAm")+ylim(0,1)+xlab("Days Post Dose 1")
}

plt(c("cg03649429","cg03736807"))
plt(c("cg05730183","cg07806979"))

order_db<-stat_biohits[rev(order(abs(stat_biohits$max_deltabeta))),]
plt(order_db[1:4,"CpG"])

order_pval<-stat_biohits[(order(abs(stat_biohits$fdr_pval))),]
plt(order_pval[1:4,"CpG"])

```


## make a summary table
```{r}
load("~/MethylationEPIC_v-1-0_B3.RData")

CpGs<-stat_biohits$CpG

annotation_stuff<-annotation_EPIC[which(annotation_EPIC$IlmnID%in%CpGs),c("IlmnID","CHR","MAPINFO","UCSC_RefGene_Name","UCSC_RefGene_Group","UCSC_CpG_Islands_Name","Relation_to_UCSC_CpG_Island","Phantom5_Enhancers","TFBS_NAME","TFBS_Evidence_Count","DNase_Hypersensitivity_NAME","DNase_Hypersensitivity_Evidence_Count")]

vaccination_stuff<-Hit_info[which(Hit_info$CpG%in%CpGs),]

gene_hit_table<-merge(vaccination_stuff, annotation_stuff, by.x="CpG", by.y="IlmnID")

gene_hit_table

gene_hit_table$Nominal_P<-round(gene_hit_table$Nominal_P,5)
gene_hit_table$fdr_pval<-round(gene_hit_table$fdr_pval,3)
gene_hit_table$max_deltabeta<-round(gene_hit_table$max_deltabeta,3)

write.csv(gene_hit_table, file="Visit_gene_table.csv")

gene_names<-unique(unlist(strsplit(as.character(gene_hit_table$UCSC_RefGene_Name),";")))

## gene specifc
gene_hit_table[grep("COL4A3BP", gene_hit_table$UCSC_RefGene_Name),]
gene_hit_table[grep("ABCG2", gene_hit_table$UCSC_RefGene_Name),]

## representative CpGS
plt(c("cg03649429","cg20474604","cg16432182","cg23118773","cg03736807","cg05730183"))

```









``` {r}
## Should you do a numeric visit as well of instead
levels(meta_vaccination$VisitNum)
meta_vaccination$visitnumeric<-meta_vaccination$VisitNum
levels(meta_vaccination$visitnumeric)<-c("","3","4","5","6","7")
meta_vaccination$visitnumeric<-as.numeric(as.character(meta_vaccination$visitnumeric))

```

